---
title: "RPlots"
author: "CSV Pitchers"
date: "November 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1.Frequency of occurrence of event type on the basis of location over the year

In this plot we are reading the data for year 2016 taking into consideration longitudes, latitudes and event types.

### Loading Data

```{r , include = TRUE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)


Dataset=read.csv('./../../RawDataset/project_data/StormEvents_details-ftp_v1.0_d2016_c20170918.csv', stringsAsFactors = TRUE)
Dataset1<-Dataset[,c(8,9,11,13,45,46,47,48)]
Dataset2<- na.omit(Dataset1)

# Insert a new Column named Region and categorize states into different Regions ("South", "West", "Midwest", "Northeast", "USA")
Dataset2$Region[Dataset2$STATE %in% c("CONNECTICUT","MAINE","MASSACHUSETTS", "NEW HAMPSHIRE","RHODE ISLAND", "VERMONT", "NEW JERSEY", "NEW YORK", "PENNSYLVANIA")] <- "Northeast"

Dataset2$Region[Dataset2$STATE%in% c("ILLINOIS", "INDIANA", "MICHIGAN", "OHIO", "WISCONSIN", "IOWA", "KANSAS", "MINNESOTA", "MISSOURI", "NEBRASKA","NORTH DAKOTA", "SOUTH DAKOTA")] <- "Midwest"

Dataset2$Region[Dataset2$STATE %in% c("DELAWARE", "FLORIDA", 
                                      "GEORGIA", "MARYLAND", 
                                      "NORTH CAROLINA", "SOUTH CAROLINA", 
                                      "VIRGINIA", "DISTRICT OF COLUMBIA", 
                                      "WEST VIRGINIA", "ALABAMA", "KENTUCKY",
                                      "MISSISSIPPI", "TENNESSEE", "ARKANSAS", 
                                      "LOUISIANA", "OKLAHOMA", "TEXAS"
)] <- "South"

Dataset2$Region[Dataset2$STATE %in% c("ARIZONA", "COLORADO", "IDAHO", 
                                      "MONTANA", "NEVADA", "NEW MEXICO", 
                                      "UTAH", "WYOMING", "ALASKA", 
                                      "CALIFORNIA", "HAWAII", "OREGON", 
                                      "WASHINGTON"
)] <- "West"

# Making Subset of Data According to the Region

south_data <-Dataset2[Dataset2$Region == 'South',]
north_data <-Dataset2[Dataset2$Region == 'Northeast',]
midwest_data <-Dataset2[Dataset2$Region == 'Midwest',]
west_data <-Dataset2[Dataset2$Region == 'West',]


# Omitting values with na values
south_data<- na.omit(south_data)
# Left With 15,778 entries
north_data<- na.omit(north_data)
# Left with 2742 entries
midwest_data<- na.omit(midwest_data)
#Left with 12,128 entries
west_data<- na.omit(west_data)
#Left with 3148 entries

#mapping data for plots
mapping_n<- north_data
mapping_n$EVENT_TYPE<- factor(north_data$EVENT_TYPE)
mapping_s<- south_data
mapping_s$EVENT_TYPE<- factor(south_data$EVENT_TYPE)
mapping_w<- west_data
mapping_w$EVENT_TYPE<- factor(west_data$EVENT_TYPE)
mapping_mw<- midwest_data
mapping_mw$EVENT_TYPE<- factor(midwest_data$EVENT_TYPE)
```

```{r, message = FALSE}
# Creating the plot under ggplot for north region
# North Data Plotting


ggplot(mapping_n,aes(STATE, EVENT_TYPE))+ geom_point(alpha=1/2, size=5, aes(color=factor(EVENT_TYPE))) + theme(axis.text.x = element_text(angle = 90, hjust =1))


# The graph shows the state on horizontal axis and events on vertical axis and the points shows states having events

# south Data Plotting

ggplot(mapping_s,aes(STATE, EVENT_TYPE)) + geom_point(alpha=1/2, size=5, aes(color=factor(EVENT_TYPE)))+theme(axis.text.x = element_text(angle = 90, hjust =1))


# West Data Plotting

ggplot(mapping_w,aes(STATE, EVENT_TYPE)) + geom_point(alpha=1/2, size=5, aes(color=factor(EVENT_TYPE)))+theme(axis.text.x = element_text(angle = 90, hjust =1))


# MidWest Data Plotting

ggplot(mapping_mw,aes(STATE, EVENT_TYPE)) + geom_point(alpha=1/2, size=5, aes(color=factor(EVENT_TYPE))) + theme(axis.text.x = element_text(angle = 90, hjust =1))

```

## Q3.Event which is more likely to occur on the basis of month (monthwise occurrence)

### Loading & Preparing Data

```{r , include = TRUE, message = FALSE,  warning = FALSE}
library(knitr)
library(ggplot2)

##Reading the file from working directory 
# Reading the Event Id,Event Type, Month and assigning it as Dataset3_month
Dataset=read.csv('./../../RawDataset/project_data/StormEvents_details-ftp_v1.0_d2016_c20170918.csv')
Dataset3_month <- Dataset[,c(8,12,13)]
# Create Month_Name as the factor.
Dataset3_month$MONTH_NAME=as.factor(Dataset3_month$MONTH_NAME)

#Showing the month with frequency of event occurance-
table_month <- table(Dataset3_month$MONTH_NAME)
View(table_month)
summary(Dataset3_month$MONTH_NAME)

#Creating Subset of data for each month-

jan_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'January',]
feb_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'February',]
mar_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'March',]
apr_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'April',]
may_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'May',]
jun_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'June',]
jul_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'July',]
aug_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'August',]
sep_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'September',]
oct_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'October',]
nov_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'November',]
dec_data <-Dataset3_month[Dataset3_month$MONTH_NAME == 'December',]

# Creating Table for jan data and assigning it to new dataframe-
table_jan <- table(jan_data$EVENT_TYPE)
data_jan<- as.data.frame(table_jan)

table_feb <- table(feb_data$EVENT_TYPE)
data_feb<- as.data.frame(table_feb)

table_mar <- table(mar_data$EVENT_TYPE)
data_mar<- as.data.frame(table_mar)

table_apr <- table(apr_data$EVENT_TYPE)
data_apr<- as.data.frame(table_apr)

table_may <- table(may_data$EVENT_TYPE)
data_may<- as.data.frame(table_may)

table_jun <- table(jun_data$EVENT_TYPE)
data_jun<- as.data.frame(table_jun)

table_jul <- table(jul_data$EVENT_TYPE)
data_jul<- as.data.frame(table_jul)

table_aug <- table(aug_data$EVENT_TYPE)
data_aug<- as.data.frame(table_aug)

table_sep <- table(sep_data$EVENT_TYPE)
data_sep<- as.data.frame(table_sep)

table_oct <- table(oct_data$EVENT_TYPE)
data_oct<- as.data.frame(table_oct)

table_nov <- table(nov_data$EVENT_TYPE)
data_nov<- as.data.frame(table_nov)

table_dec <- table(dec_data$EVENT_TYPE)
data_dec<- as.data.frame(table_dec)

```

```{r, message = FALSE}
# Creating plot-

p1 <-  ggplot(data_jan, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "January Data Plot",x="Events", y="Number of Events")
p1

p2 <-  ggplot(data_feb, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "Feb Data Plot",x="Events", y="Number of Events")
p2

p3 <-  ggplot(data_mar, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "March Data Plot",x="Events", y="Number of Events")
p3
p4 <-  ggplot(data_apr, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "April Data Plot",x="Events", y="Number of Events")
p4
p5 <-  ggplot(data_may, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "May Data Plot",x="Events", y="Number of Events")
p5
p6 <-  ggplot(data_jun, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "June Data Plot",x="Events", y="Number of Events")
p6

p7 <-  ggplot(data_jul, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "July Data Plot",x="Events", y="Number of Events")
p7

p8 <-  ggplot(data_aug, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "August Data Plot",x="Events", y="Number of Events")
p8

p9 <-  ggplot(data_sep, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "September Data Plot",x="Events", y="Number of Events")
p9

p10 <-  ggplot(data_oct, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "October Data Plot",x="Events", y="Number of Events")
p10

p11 <-  ggplot(data_nov, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "November Data Plot",x="Events", y="Number of Events")
p11

p12 <-  ggplot(data_dec, aes(x=Var1, y = Freq))+geom_col()+ theme(axis.text.x = element_text(angle = 90, hjust =1))+labs(title = "December Data Plot",x="Events", y="Number of Events")
p12
```
## Q4.  What is estimated amount of damage (damage property & damage crops) caused by each event? which event is most likely to cause maximum damage?

### Loading & Preparing Data

```{r , include = TRUE, message = FALSE,  warning = FALSE}

library(knitr)
library(dplyr)
library(ggplot2)
library(scales)

#Read  the dataset of 2016
Dataset1=read.csv('./../../RawDataset/project_data/StormEvents_details-ftp_v1.0_d2016_c20170918.csv',stringsAsFactors = F)

#Extract only Event Id,Event Type, Month, damage property, damage crops and assigning it as Dataset4_damage

Dataset4_damage <- Dataset1[,c(8,12,13,25,26)]

# Multiply the values of amount in damage_property and damage_crops containing M in data by 1000 to turn into K and remove M from data
# Remove K from data

Dataset4_update <- Dataset4_damage

index1 <- (substr(Dataset4_update$DAMAGE_PROPERTY,nchar(Dataset4_update$DAMAGE_PROPERTY),nchar(Dataset4_update$DAMAGE_PROPERTY)))=="M"
index2 <- (substr(Dataset4_update$DAMAGE_CROPS,nchar(Dataset4_update$DAMAGE_CROPS),nchar(Dataset4_update$DAMAGE_CROPS)))=="M"
index3 <- (substr(Dataset4_update$DAMAGE_PROPERTY,nchar(Dataset4_update$DAMAGE_PROPERTY),nchar(Dataset4_update$DAMAGE_PROPERTY)))=="K"
index4 <- (substr(Dataset4_update$DAMAGE_CROPS,nchar(Dataset4_update$DAMAGE_CROPS),nchar(Dataset4_update$DAMAGE_CROPS)))=="K"


Dataset4_update$DAMAGE_PROPERTY[index1] <- 
  as.numeric(substr(Dataset4_update$DAMAGE_PROPERTY[index1],1,nchar(Dataset4_update$DAMAGE_PROPERTY[index1])-1))*1000  

Dataset4_update$DAMAGE_CROPS[index2] <- 
  as.numeric(substr(Dataset4_update$DAMAGE_CROPS[index2],1,nchar(Dataset4_update$DAMAGE_CROPS[index2])-1))*1000  

Dataset4_update$DAMAGE_PROPERTY[index3] <- 
  as.numeric(substr(Dataset4_update$DAMAGE_PROPERTY[index3],1,nchar(Dataset4_update$DAMAGE_PROPERTY[index3])-1))

Dataset4_update$DAMAGE_CROPS[index4] <- 
  as.numeric(substr(Dataset4_update$DAMAGE_CROPS[index4],1,nchar(Dataset4_update$DAMAGE_CROPS[index4])-1)) 

#changing fields into numeric
Dataset4_update$DAMAGE_PROPERTY <- as.numeric(as.character(Dataset4_update$DAMAGE_PROPERTY))
Dataset4_update$DAMAGE_CROPS <- as.numeric(as.character(Dataset4_update$DAMAGE_CROPS))

#creating dataset removing data having NA in value
Dataset4_update_rm<- na.omit(Dataset4_update)

#creating dataset replacing data having NA with 0
Dataset4_update_na<- Dataset4_update
Dataset4_update_na$DAMAGE_PROPERTY[is.na(Dataset4_update_na$DAMAGE_PROPERTY) ] <- 0

Dataset4_update_na$DAMAGE_CROPS[is.na(Dataset4_update_na$DAMAGE_CROPS) ] <- 0

#Add column which contains the sum of amount of damage property and damage crops
Dataset4_na_addcol <- Dataset4_update_na
Dataset4_na_addcol$DAMAGE_OVERALL <- (Dataset4_na_addcol$DAMAGE_PROPERTY + Dataset4_na_addcol$DAMAGE_CROPS)

```

```{r, message = FALSE}
blue.bold.italic.text <- element_text(face = "bold.italic", color = "blue")


#plot for damage property
p <-  ggplot(Dataset4_na_addcol, aes(x=EVENT_TYPE, y = DAMAGE_PROPERTY))+geom_line()
 p+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Damage of property by events",
        x="Event Types", y="Amount of Damage $ (in K)")

#plot for damage crop

 c <-  ggplot(Dataset4_na_addcol, aes(x=EVENT_TYPE, y = DAMAGE_CROPS))+geom_line()
 c+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Damage of crops by events",
        x="Event Types", y="Amount of Damage $ (in K)")

 #plot for overall damage 
 o <-  ggplot(Dataset4_na_addcol, aes(x=EVENT_TYPE, y = DAMAGE_OVERALL))+geom_line()
 o+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Damage of crops by events",
        x="Event Types", y="Amount of Damage $ (in K)")
```

## Q5. Which event is the most severe taking into consideration both fatalities and damage?

### Loading & Preparing Data

```{r , include = TRUE, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(scales)

#Read  the dataset of 2016

Dataset1=read.csv('./../../RawDataset/project_data/StormEvents_details-ftp_v1.0_d2016_c20170918.csv',stringsAsFactors = F)

#Extract only Event Id,Month name, Event Type, damage property, Injuries indirect, injuries direct, deaths indirect, deaths direct and assigning it as Dataset5_severity

Dataset5_severity <- Dataset1[,c(8,12,13,25,21,22,24,23)]

# Multiply the values of amount in damage_property containing M in data by 1000 to turn into K and remove M from data
# Remove K from data

Dataset5_updateSeverity <- Dataset5_severity
index1 <- (substr(Dataset5_updateSeverity$DAMAGE_PROPERTY,nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY),nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY)))=="M"

index2 <- (substr(Dataset5_updateSeverity$DAMAGE_PROPERTY,nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY),nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY)))=="K"


Dataset5_updateSeverity$DAMAGE_PROPERTY[index1] <- 
  as.numeric(substr(Dataset5_updateSeverity$DAMAGE_PROPERTY[index1],1,nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY[index1])-1))*1000  

 
Dataset5_updateSeverity$DAMAGE_PROPERTY[index2] <- 
  as.numeric(substr(Dataset5_updateSeverity$DAMAGE_PROPERTY[index2],1,nchar(Dataset5_updateSeverity$DAMAGE_PROPERTY[index2])-1))

#changing fields into numeric
Dataset5_updateSeverity$DAMAGE_PROPERTY <- as.numeric(as.character(Dataset5_updateSeverity$DAMAGE_PROPERTY))

Dataset5_updateSeverity$DAMAGE_PROPERTY[is.na(Dataset5_updateSeverity$DAMAGE_PROPERTY) ] <- 0
```

```{r, message = FALSE}
#plot for INJURIES_INDIRECT

p <-  ggplot(Dataset5_updateSeverity, aes(x=EVENT_TYPE, y = INJURIES_INDIRECT))+geom_line()
 p+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Indirect injuries by events",
        x="Event Types", y="Amount of INJURIES_INDIRECT $ (in K)")		

#plot for INJURIES_DIRECT

p <-  ggplot(Dataset5_updateSeverity, aes(x=EVENT_TYPE, y = INJURIES_DIRECT))+geom_line()
 p+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Direct injuries by events",
        x="Event Types", y="Amount of INJURIES_DIRECT $ (in K)")
		
#plot for DEATHS_DIRECT

p <-  ggplot(Dataset5_updateSeverity, aes(x=EVENT_TYPE, y = DEATHS_DIRECT))+geom_line()
 p+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Direct Deaths by events",
        x="Event Types", y="Amount of DEATHS_DIRECT $ (in K)")	

#plot for DEATHS_INDIRECT

p <-  ggplot(Dataset5_updateSeverity, aes(x=EVENT_TYPE, y = DEATHS_INDIRECT))+geom_line()
 p+ theme(axis.text.x = element_text(angle = 90, hjust =1),
          title = blue.bold.italic.text, axis.title = blue.bold.italic.text)+
   labs(title = "Indirect Deaths by events",
        x="Event Types", y="Amount of DEATHS_INDIRECT $ (in K)")
 ```
